\documentclass[ProjectReport]{subfiles}

\begin{document}

\section{Future Work}

The initial extension is to finish the proof that the Dedekind reals are a complete Archimedean ordered field. This will serve as a basis for a large number of applications. The short term plan follows.

\begin{itemize}
    \item Prove the continuous extension theorem for metric spaces. It would be very useful to prove this directly, as it allows us to extend any continuous rational function to reals, and it is known that computable functions are continuous \cite{81018}. Failing this, we try and prove that Bishop continuous rational functions can be extended, which should be easier but is less general.
    \item If all attempts to prove extension theorems fail, then instead we can directly implement multiplication of reals. It is known to be egregious, and may be lengthy but should be possible to prove. 
    \item Conclude by proving the field axioms for Dedekind reals, and prove that the reals are Dedekind complete. 
\end{itemize}

With this formalisation in place, we look to the applications of the library.

\subsection{Exact Computation}

The thesis of Booij \cite{booij_2021} describes how the reals may be equipped with locators to observe information. It is shown that ``we can compute arbitrarily precise approximations to reals with locator", but stops short of doing so. Hence, with this development of the reals, by equipping them with locators we can work on providing verified exact computation with real numbers.

Due to various choices of representation, my code does not have any claim to efficiency of computation. I have chosen representations of numbers that are conducive to writing \textit{proofs} efficiently. As an explicit example, there would be an added layer of difficulty to use a binary representation of natural numbers to formalise the reals. Using binary induction is possible, but adds complexity to proofs that are much simpler using natural induction. 

For the purpose of exact computation it would be prudent to have isomorphic representations which are more suitable for the purposes of efficient computation. A nice extension of this project would be to write these efficient representations, showing that they are isomorphic to my definitions, and provide an interface for exact computation of real numbers. 

\subsection{HoTT Book}

Having used the HoTT Book as guidance for some of the proofs in this development, it would be useful to continue the formalisation that follows in the book. In particular, I am interested in formalising the Cauchy reals, and show that they can be embedded in the Dedekind Reals. I am also interested in the role of the Dedekind reals in homotopy type theory. One investigation is into the link between elements of synthetic homotopy theory developed using the Dedekind reals, and constructive analysis. 

\subsection{Formalisation}

Formalisation is an important part of modern mathematics. Verification of proofs provides guarantees of correctness, and in particular proofs developed in a constructive setting allows us to extract useful programs. The reals as a foundations are conducive to extensions in many studies in mathematics. I am personally interested in formalising concepts in real and complex analysis, in numerical analysis, and the topology of real numbers. 

\end{document}